---
title: "condrift: Conditional drift in an admixture graph"
output:
  html_notebook
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

suppressPackageStartupMessages(library(ggplot2, quietly = TRUE))
suppressPackageStartupMessages(library(admixturegraph, quietly = TRUE))
suppressPackageStartupMessages(library(condrift, quietly = TRUE))
```

We can consider an admixture graph like the following:

```{r, echo=FALSE}
leaves <- c("D", "Alt", "V", "EU", "EA", "Af", "C")
inner_nodes <- c("R", "X", "Y", "Z", "W", "x", "d", "w", "U", "v", "xx")

edges <- parent_edges(c(
  edge("C", "R"),
  edge("X", "R"),
  edge("Y", "X"),
  edge("x", "X"),
  edge("Z", "Y"),
  admixture_edge("d", "x", "Z", "a"),
  edge("D", "d"),
  edge("W", "Z"),
  edge("Alt", "W"),
  edge("w", "W"),
  edge("V", "w"),
  edge("U", "Y"),
  admixture_edge("v", "w", "U", "b"),
  edge("xx", "v"),
  edge("EA", "xx"),
  edge("EU", "xx"),
  edge("Af", "U")
))

graph <- agraph(leaves, inner_nodes, edges)
plot(graph, show_admixture_labels = TRUE)
```

The graph topology together with the edge lengths and admixture proportions define a covariance matrix $\Sigma$ where entry $\sigma_{ij}$ is given by the $F_3(r;i,j)$ statistics where $r$ is an arbitrary leaf we choose to root our covariance matrix in (see Felsenstein 2004 chapter 23). This statistics we can both infer from the data and extract from a fitted graph. I haven't fitted the graph here, so the covariance matrix shown below is completely arbitrary, but it is extracted from the graph where I've used $C$ as root.

```{r, echo=FALSE}
edge_lengths <- list2env(list(
  a = 0.05, b = 0.02, edge_d_D = 0.008, edge_Z_d = 0.005, edge_W_Alt = 0.001, 
  edge_Z_W = 0.001, edge_X_x = 0.1, edge_X_Y = 0.002, edge_Y_Z = 0.2,
  edge_x_d = 0.002, edge_w_V = 0.002, edge_W_w = 0.002, edge_U_v = 0.003,
  edge_v_xx = 0.01, edge_xx_EU = 0.001, edge_Y_U = 0.01, edge_w_v = 0.003, 
  edge_xx_EA = 0.0003, edge_U_Af = 0.004, edge_R_C = 0.2, edge_R_X = 0.004))

Sigma <- extract_covariance_matrix(graph, edge_lengths, "C")
Sigma
```

```{r, echo=FALSE}
plot.drift_model <- function(x, ..., indices = c(1, 2)) {
  i <- indices[1]
  j <- indices[2]
  if (is.character(i))
    i <- which(i == rownames(x$Sigma))
  if (is.character(j))
    j <- which(j == rownames(x$Sigma))
  
  mu <- x$mu[c(i,j)]
  Sigma <- x$Sigma[c(i,j), c(i,j)]
  
  range1 <- range2 <- seq(0, 1, length.out = 200)
  data.grid <- expand.grid(f1 = range1, f2 = range2)
  q.samp <- cbind(data.grid, 
                  prob = mvtnorm::dmvnorm(data.grid, mean = mu, sigma = Sigma))

  ggplot(q.samp, aes(x = f1, y = f2, z = prob)) + 
    geom_contour() +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1), ratio = 1)
}
```

Since the graph gives us a multivariate Normal distribution of allele frequencies, we can relatively easy compute conditional probabilities, capturing what the correlation of two populations should be conditional on another, for example.

```{r}
mu <- rep(0.5, length(graph$leaves) - 1) # arbitrary choice...
names(mu) <- graph$leaves[graph$leaves != "C"]
model <- joint_drift_model(mu, Sigma)

plot_conditional_distribution <- function(a, b, x, f) {
  plot(condition(model, index = x, freq = f), indices = c(a, b)) +
    xlab(a) + ylab(b)
}

plot_conditional_distribution("D", "V", "Af", f = 0.01)
plot_conditional_distribution("D", "V", "Af", f = 0.3)
plot_conditional_distribution("D", "V", "Af", f = 0.7)
plot_conditional_distribution("D", "V", "Af", f = 0.9)

plot_conditional_distribution("EU", "EA", "Af", f = 0.1)
plot_conditional_distribution("EU", "EA", "Af", f = 0.5)
plot_conditional_distribution("EU", "EA", "Af", f = 0.9)
```

```{r}
expected_F4 <- function(model, a, b, c, d) {
  # E[(a-b)(c-d)] = E[ac]-E[ad]-E[bc]+E[bd]
  # E[xy] = E[x]E[y] + Cov(x,y)
  Eac <- model$mu[a] * model$mu[c] + model$Sigma[a, c]
  Ead <- model$mu[a] * model$mu[d] + model$Sigma[a, d]
  Ebc <- model$mu[b] * model$mu[c] + model$Sigma[b, c]
  Ebd <- model$mu[b] * model$mu[d] + model$Sigma[b, d]
  as.vector(Eac - Ead - Ebc + Ebd)
}

conditional_expected_F4 <- function(model, a, b, c, d, x) {
  cond <- function(f) {
    conditional_model <- condition(model, index = x, freq = f)
    expected_F4(conditional_model, a, b, c, d)  
  }
  Vectorize(cond)
}
```

```{r}
freqs <- seq(0, 1, length.out = 200)

cond_F4 <- conditional_expected_F4(model, "D", "V", "Af", "Alt", "EU")
plot(freqs, cond_F4(freqs), type = 'l', xlab = "EU allele freq", 
     ylab = quote(F[4](D, V, Af, Alt)))

cond_F4 <- conditional_expected_F4(model, "EU", "EA", "V", "Alt", "Af")
plot(freqs, cond_F4(freqs), type = 'l', xlab = "Af allele freq", 
     ylab = quote(F[4](EU, EA, V, Alt)))
```
